---
title: "Final Project: Predictive Modeling of Mice Neural Activity in Response to Visual Stimuli"
author: "Tong Zhu"
date: "2024-03-18"
output: html_document
---

```{r echo=FALSE,include=FALSE}
#set up for the project
#load library
library(ggplot2)
```

# Abstract
This report started with an analysis of mice's neural response to different 
visual stimuli based on the dataset by Steinmetz et al. (2019). Then it aims 
to integrate mice's decision-making process and train a model to predict the 
outcome of mice's decision given a random dataset selected from the same 
experiment.  Through the integration of data, this report tries to identify the 
distribution of the data across sessions and trials, exploring the variation in 
neural responses and trial results(correctness of response). After finding the 
shared neural patterns across sessions, the predicted model is trained and can 
be used to evaluate test sets randomly selected from distinct sessions. The 
outcomes of this research enhanced our understanding of the relationship between 
decision-making and neural activity.

# I. Introduction
The experiments were performed on a total of 10 mice over 39 
sessions, this report uses 4 mice over 18 sessions as a dataset, focusing on 
the contrast levels of stimuli presented and neural activity from each trial.The 
prediction model will be trained and selected by the performance of result. The 
test data set will be evaluated on two test sets of 100 trials randomly selected 
from Session 1 and Session 18, respectively. 

# II. Exploratory analysis

### Describe the data structures across sessions
After processing through each trial from each session, we can observe that there
are 6 variables for each trial, namely:
*feedback_type*: type of the feedback, 1 for success and -1 for failure,
*contrast_left*: contrast of the left stimulus(0, 0.25, 0.5, 1),
*contrast_right*: contrast of the right stimulus(0, 0.25, 0.5, 1),
*time*: centers of the time bins for spks,
*spks*: numbers of spikes of neurons in the visual cortex in time bins,
*brain_area*: area of the brain where each neuron lives.

There are 4 distinct mice recorded in the dataset, named Cori(session1-3), 
Forssmann(session4-7), Hench(session8-11), and Lederberg(session12-18).
A bar plot of overall correctness across session is as follows, from the bar 
plot, we can observe that some of the session have higher corretness,we can do 
further integration on these trials.



```{r echo=FALSE,warning=FALSE,message=TRUE}
#load data
session <- list()
for (i in 1:18) 
  {
  filepath <- file.path("E:/study/DAVIS/Winter2024/sta141A/final project/Data/session", paste('session', i, '.rds', sep = ''))
  session[[i]] <- readRDS(filepath)
}

#create a data frame for session summaries
sessions_summary <- data.frame(
  Session = integer(),
  Number_of_Trials = integer(),
  Proportion_Success = numeric(),
  Proportion_Failure = numeric(),
  Mean_Stimuli_Left = numeric(),
  Mean_Stimuli_Right = numeric(),
  stringsAsFactors = FALSE
)

#go through each session
for (i in seq_along(session)) {
  current_session <- session[[i]]
  
  #create vectors to store success and failure data
  success_per_trial <- numeric()
  failure_per_trial <- numeric()
  stimuli_left <- numeric()
  stimuli_right <- numeric()

  #go through each trial within session
  for (j in seq_along(current_session$spks)) {
    #append success and failure data
    success_per_trial[j] <- mean(current_session$feedback_type[[j]] == 1)
    failure_per_trial[j] <- mean(current_session$feedback_type[[j]] == -1)
    
    # Combine all stimulus levels across all trials
    stimuli_left <- c(stimuli_left, current_session$contrast_left[[j]])
    stimuli_right <- c(stimuli_right, current_session$contrast_right[[j]])
  }

  #summary for each session
  session_summary <- data.frame(
    Session = i,
    Number_of_Trials = length(current_session$spks),
    Proportion_Success = mean(success_per_trial),
    Proportion_Failure = mean(failure_per_trial),
    Mean_Stimuli_Left = mean(stimuli_left, na.rm = TRUE),
    Mean_Stimuli_Right = mean(stimuli_right, na.rm = TRUE)
  )
  
  #add current session summery to overall summary
  sessions_summary <- rbind(sessions_summary, session_summary)
}
```

```{r echo=FALSE,warning=FALSE,message=TRUE}
#bar plot correctness vs session
barplot(sessions_summary$Proportion_Success,
        names.arg=paste('Session', sessions_summary$Session),
        xlab='Session',
        ylab='Correctness (Proportion Success)',
        main='Correctness by Session',
        col='lightblue',
        ylim=c(0, 1))

#add success rate above each bar
text(x = barplot(sessions_summary$Proportion_Success, plot=FALSE), 
     y = sessions_summary$Proportion_Success + 0.02, 
     label = round(sessions_summary$Proportion_Success, 2),
     cex = 0.7,
     pos = 3)
```

From the plot, we can see that all 4 mice is doing better than expected(0.5), 
Lederberg has an outstanding performance comparing to other mice.Session 11,13,
17,18 have higher correctness rate, these session may worth further analyze. 


```{r echo=FALSE,include=TRUE}
#extract all brain area data
all_brain_areas <- unlist(lapply(session, function(s) unlist(s$brain_area)))

#count the frequency of each brain area
brain_area_counts <- table(all_brain_areas)

#sort and find the most common brain areas
sorted_brain_area_counts <- sort(brain_area_counts, decreasing = TRUE)

#bar plot
top_n <- 10 
barplot(sorted_brain_area_counts[1:top_n], las=2, main="Most Common Used Brain Areas",
        xlab="Brain Areas", ylab="Counts", col="lightblue", cex.names=0.7)

```

### Neuron activity
We want to know about neuron activity in each trial so we extract spikes and 
time from single trials and create raster plot.

```{r echo=FALSE,include=TRUE}
plot_neural_activity <- function(spikes, time, trial_index) {
  trial_spikes <- spikes[[trial_index]]
  num_neurons <- nrow(trial_spikes)
  
  #extract the time vector
  trial_time <- time[[trial_index]]
  
  #raster plot
  plot(NULL, NULL, type = 'n', xlim = c(min(trial_time), max(trial_time)), ylim = c(0, num_neurons),
       xlab = 'Time (s)', ylab = 'Neuron Index', main = paste('Neural Activity'))
  
  #plot spikes for each neuron
  for (neuron_index in 1:num_neurons) {
    spikes_times <- trial_time[trial_spikes[neuron_index, ] > 0]
    if (length(spikes_times) > 0) {
      points(spikes_times, rep(neuron_index, length(spikes_times)), pch = '|')
    }
  }
}

plot_neural_activity(session[[7]]$spks, session[[7]]$time, 20)
plot_neural_activity(session[[18]]$spks, session[[18]]$time, 1)
```

From above plot, session 7,trial 20 and session 18 trial 1 is randomly choose 
to see the raster plot of neuron and time. The same function can be called for 
all trials across sessions. From the plot, we can see directly how the neurons 
active over time during the trial. 

### Explore changes over trials
We use heat map as the method to explore changes over trials with in sessions. 
18 heat maps are created corresponding each session.Each row in a heat map 
corresponds to a single trial, while each column represents a time bin within 
that trial. The color intensity reflects the normalized spike rate, with warmer 
colors indicating higher levels of neural firing.

```{r echo=FALSE,include=FALSE}
plot_activity_heatmap <- function(spks, num_trials_to_plot=NULL) {
  if (is.null(num_trials_to_plot)) {
    num_trials_to_plot <- length(spks)
  }

  #empty matrix to store aggregated spike data 
  spike_matrix <- matrix(nrow = num_trials_to_plot, ncol = 40)
  
  #loop all trials and record in the spike matrix
  for (i in 1:num_trials_to_plot) {
    spike_matrix[i, ] <- colSums(spks[[i]])
  }
  spike_matrix <- spike_matrix / max(spike_matrix)
  nr <- nrow(spike_matrix)
  nc <- ncol(spike_matrix)

  #plot the heat map
  image(x = 1:nc, y = 1:nr, z = t(spike_matrix)[, nr:1], 
        xlab = 'Time Bins', ylab = 'Trials', main = 'Neural Activity Across Trials',
        axes = FALSE)
}

# Plot the heat map for every session
for (i in (1:18))
{
  plot_activity_heatmap(session[[i]]$spks)
}

```
In examining all heat maps, we notice several recurring patterns: There are 
periods of heightened activity that appear consistently across trials, 
indicating that the neurons are responsive to the visual task given.
Session 17 has the highest correctness rate from the previous data exploration, 
and session 11 has a relatively high correctness rate but we think the pattern 
is worth analyze, hence we attach the heat map of session 11 and session 17 as 
follows, we can see a obvious pattern that a large proportion of trials have a 
similarity in time bins which indicates consistent neural responses to the 
experimental conditions.

```{r echo=FALSE,include=TRUE}
plot_activity_heatmap(session[[11]]$spks)
plot_activity_heatmap(session[[17]]$spks)
```

### Explore homogeneity and heterogeneity across sessions and mice.
We are trying to explore homogeneity and heterogeneity across sessions and mice
by calculating the variance of total spikes within each session and mean total 
spikes per session for each mouse. The plots are as follows:

```{r echo=FALSE,include=TRUE}
#get the total spikes per trial/per session
total_spikes_per_session <- lapply(session, function(s) {
  unlist(lapply(s$spks, function(trial) sum(trial)))
})

#variance of total spikes for each session
variance_per_session <- sapply(total_spikes_per_session, var)

#plot the variance
barplot(variance_per_session, names.arg=paste('Session', seq_along(variance_per_session)),
        xlab='Session', ylab='Variance of Total Spikes', main='Variance of Total Spikes Within Each Session')
```


A higher variance in total spike count indicates a substantial variation in the 
number of spikes from trial to trial within those sessions. This may caused by 
control factors, differences in mouse, and variability in neural responses. A 
lower variance suggesting a more consistent neural response across trials.

```{r echo=FALSE,include=TRUE}
#vector with the mouse name
mouse_names <- c(rep("Cori", 3), rep("Forssmann", 4), rep("Hench", 4), rep("Lederberg", 7))

#vector total spikes
total_spikes_vector <- unlist(total_spikes_per_session)

#mouse name for each trial
mouse_factors <- rep(mouse_names, sapply(total_spikes_per_session, length))

#mean total spikes per session for each mouse
mean_spikes_per_mouse <- tapply(total_spikes_vector, mouse_factors, mean)

#plot
barplot(mean_spikes_per_mouse, names.arg = names(mean_spikes_per_mouse),
        xlab = 'Mouse', ylab = 'Mean Total Spikes', main = 'Mean Total Spikes Per Session for Each Mouse')

```

The mean total spikes appear to be relatively consistent across mice, with only 
slight variations between them.Forssmann and Lederberg has a slightly lower mean 
total spikes.This may be caused by individual differences between mice, such as 
decision-making processing, and learning speed throughout the experiment.

# III. Data integration
Extract the shared patters across sessions and address the differences between 
sessions.The training data for performing PCA is previewed as follows. 
```{r echo=FALSE,include=TRUE}
#prepare training data
training_data_list <- lapply(1:length(session), function(i) {
  data <- session[[i]]
  mouse_name_vector <- rep(data$mouse_name, times = length(data$feedback_type))
  session_id_vector <- rep(i, times = length(data$feedback_type))
  contrast_left_vector <- data$contrast_left
  contrast_right_vector <- data$contrast_right
  contrast_diff_vector <- contrast_left_vector - contrast_right_vector
  feedback_type_vector <- data$feedback_type
  
  #total spikes per trial
  total_spikes_vector <- sapply(data$spks, function(trial_matrix) sum(trial_matrix))

  #data frame
  data_frame <- data.frame(
    MouseName = mouse_name_vector,
    SessionID = session_id_vector,
    ContrastLeft = contrast_left_vector,
    ContrastRight = contrast_right_vector,
    ContrastDiff = contrast_diff_vector,
    TotalSpikes = total_spikes_vector,
    FeedbackType = feedback_type_vector
  )
  return(data_frame)
})

#combine train data
training_tibble <- do.call(rbind, training_data_list)

#head of training data
head(training_tibble)
```
Next, we perform the PCA to see if what pattern is shared across sessions. 




# IV. Predictive modeling
```{r echo=FALSE,include=TRUE}
#read test data
set.seed(123)
test1 <- readRDS("E:/study/DAVIS/Winter2024/sta141A/final project/Data/test_data/test1.rds")
test2 <- readRDS("E:/study/DAVIS/Winter2024/sta141A/final project/Data/test_data/test2.rds")


#predictions on the test sets
#predictions_test1 <- predict(model, test1)
#predictions_test2 <- predict(model, test2)

#evaluate
#confusionMatrix(predictions_test1, test1$FeedbackType)
#confusionMatrix(predictions_test2, test2$FeedbackType)
```

# V. Prediction performance on the test sets
fail in establish prediction model leads to fail in predict of test sets.
# VI. Discussion
# Reference
Steinmetz, N.A., Zatka-Haas, P., Carandini, M. et al. Distributed coding of choice, action and engagement across the mouse brain. Nature 576, 266–273 (2019). https://doi.org/10.1038/s41586-019-1787-x

# Use of online source
This report uses chatgpt for brain storming and grammar check. Use Stack 
overflow to check use of r commands and use of prediction models. 


